Measuring Hash Trustworthiness via Collision Utility Metrics: Logical Cryptanalysis of MD4

### 2017-12-02 dsn-2018/bare_conf.tex
Added collisions to paper
Signed-off-by: Alexander Scheel <alexander.m.scheel@gmail.com>
Updates to introduction, related work
Signed-off-by: Alexander Scheel <alexander.m.scheel@gmail.com>

Cryptographic hash functions form the core of many protocols. With widespread
use as file integrity checks to verify long term storage of data, as cache
invalidation techniques, and as a building block in network protocols such as
Kerberos and TLS, this class of functions necessarily has strong guarantees
about properties of its members. There are three basic properties hash
functions must have to be considered cryptographically secure:
\begin{itemize}
    \item \textbf{Preimage Resistance}: It should be computationally hard to find
        the inverse of a hash function.
    \item \textbf{Second Preimage Resistance}: Given an input block, it should be
        computationally hard to find a second block which hashes to the
        same value as the first block.
    \item \textbf{Collision Resistance}: It should be computationally hard to find two
        blocks which hash to the same value.
\end{itemize}
Note that a second preimage is necessarily a collision and is also a stronger
result in practice.

From an attacker's perspective, finding a preimage or second preimage has been
difficult. To the author's knowledge, the current bound for a preimage attack
in MD5 is $2^{123.4}$ by Y. Sasaki and K. Aoki \cite{SasakiPreimage}. Further,
while J. Kelsey and B. Schneier proposed a method for finding second preimages
in less than $2^n$, we note that this requires rather long messages
\cite{SchneierSecondPreimage}. However, collision attacks are well within reach
for adversaries, and well past only theoretical attacks. The work of
H. Dobbertin \cite{Dobbertin1998}, X. Wang \cite{cryptoeprint:2004:199},
M. Schl{\"a}ffer \cite{Schlaffer2006} and others demonstrate the ease with which
collisions can be found.

@@

We seek to address the problem of evaluating the use cases for a collision,
to determine whether or not a collision affects a particular system. This
paper contributes several novel techniques to logical cryptanalysis and uses
these techniques to derive metrics of the trustworthiness of hash functions.
We claim the following results are novel in the field:
\begin{itemize}
    \item We define the neighborhood of a class of collisions and show that
        it is frequently non-empty.
    \item We discuss techniques for finding useful collisions within a
        class of collisions.
    \item We propose additional techniques for building new classes of
        collisions and show that collisions in MD4 can be found inductively.
    \item We propose five universal metrics for evaluating the utility of
        a class of collisions.
\end{itemize}

\textbf{TODO - Transition}

We propose that collision classes of high utility are closer to a second
preimage attack in the amount of flexibility they provide to an attacker,
whereas classes of low utility may not affect all systems which use a
particular hash function. Furthermore, we demonstrate new techniques for
working with collisions under a logical cryptanalysis framework. Throughout
this paper, we make use of the MD4 hash function due to the ease of generating
new classes of collisions.

\textbf{TODO - Finish Organization}

The remainder of this paper is organized as follows. In Section
\ref{Sec:Related} we discuss previous results and show how our research
expands upon prior work. In Section \ref{Sec:Terminology}, we discuss the
terminology and notations we use throughout the remainder of the paper.
In Section \ref{Sec:Intuition}, we give the intuition between measuring
the utility of a collision and motivate why it is useful. In Section
\ref{Sec:Logical}, we present new techniques in logical cryptanalysis to
analyze collisions in hash functions.

@@

\textbf{Broad context... intersection... subsection introductions.}

This paper draws upon the public cryptanalysis of MD4, the wide availability of
high quality SAT solvers, and previous work in logical cryptanalysis.

@@

\textbf{TODO - expand?}

Differential Cryptanalysis has been the major technique behind the discovery
of collisions in cryptographic hash functions. Its importance can be seen
everywhere from X. Wang's attacks on MD4 \cite{Wang2005} to more recent
cryptanalysis of general purpose hash functions such as Murmur3 by
J. Aumasson, D. Bernstein, and M. Bo{\ss}let \cite{murmur3DC}. However, while
differential cryptanalysis plays an important role, techniques for automated
analysis of hash functions are still an emerging area. In
\cite{cryptoeprint:2004:199}, X. Wang claimed to have found the MD4 collision
``with hand calculation'', though in \cite{Wang2005}, she specifies message
modification techniques which can be implemented in code.

Our work removes the need for finding message modification techniques by
specifying the differential path as part of the SAT formula and letting the
solver find a pair of satisfying messages which follow the differential path
and produce a collision. Further constraints can then be placed upon this
model, such as a chosen prefix or desired start state. We believe performing
these attacks by hand to be difficult, and note that message modification
techniques may affect the chosen prefix.

@@

\textbf{TODO - expand?}

Logical Cryptanalysis and its encoding as SAT likely started under the work of
F. Massacci in 1999 with his paper titled ``Using walk-SAT and rel-sat for
cryptographic key search''. \cite{Massacci:1999:UWR:1624218.1624261}. However,
much of the early work by F. Massacci was focused on symmetric and asymmetric
ciphers (in \cite{Massacci:1999:UWR:1624218.1624261},
\cite{MASSACCI2000LogicalCA}, and \cite{FIORINI2003101}). It wasn't until the
work of D. Jovanovi{\'{c}} and P. Jani{\v{c}}i{\'{c}} in \cite{LogicalAnalysis}
that exploring collisions in the context of SAT was introduced, and until
the work of I. Mironov and L. Zhang in \cite{Mironov2006} that this was
studied for a X. Wang's collision class.

We make use of the \textit{bc2cnf} utility by T. Junttila (available here
\cite{circuits}) as the basis for our models. This allows us to encode hash
functions as generic structures in the circuit description language and
ease the algorithmic creation of models which build on top of them. We then
use CryptoMiniSat5 by M. Soos (\cite{CryptoMiniSat5}) to run the models and
check for a satisfying witness, often an example collision.

With the exception of the work by I. Mironov, the authors find that much of
the previous work in the field has been focused on evaluating the performance
of various SAT solvers. We deviate from this by instead focusing on evaluating
the properties of the hash function--not the SAT solver. We have developed new
techniques for using a SAT solver to inductively build full collisions in MD4
from collisions in reduced-round MD4 and to build addition examples of
collision classes from a single instance of a collision in MD4.

@@

\begin{itemize}
    \item $C_{Wang}$, to be the collision class introduced by X. Wang
        \textit{et al.} in \cite{Wang2005}.
    \item $C_{Sasaki}$, to be the collision class introduced by Y. Sasaki
        \textit{et al.} in \cite{Sasaki2007}.
    \item $C_{Dobbertin}$, to be the collision class introduced by
        H. Dobbertin in \cite{Dobbertin1998}.
    \item $C_{Kasselman}$, to be the collision class introduced by
        P. Kasselman in \cite{KasselmanMD4}.
    \item $C_{Schlaffer}$, to be the collision class introduced by
        M. Schl{\"a}ffer in \cite{Schlaffer2006}.
\end{itemize}

@@

>>>>

Cryptographic hash functions form the core of many protocols. With widespread
use of hash functions as file integrity checks to verify long term storage of
data, as cache invalidation techniques, and as a building block in network
protocols such as Kerberos and TLS, this class of functions necessarily has
strong guarantees about properties of its members. There are three properties
hash functions must have to be considered cryptographically secure:
\begin{itemize}
    \item \textbf{Preimage Resistance}: It should be computationally hard to find
        the inverse of a hash function.
    \item \textbf{Second Preimage Resistance}: Given an input block, it should be
        computationally hard to find a second block which hashes to the
        same value as the first block.
    \item \textbf{Collision Resistance}: It should be computationally hard to find two
        blocks which hash to the same value.
\end{itemize}
Note that while a second preimage is necessarily a collision, it is also a
stronger result with higher use for an attacker.

From an attacker's perspective, finding a preimage or second preimage has been
difficult: the current published bound for a preimage attack in MD5 is a cost
of $2^{123.4}$ by Y. Sasaki and K. Aoki \cite{SasakiPreimage}. Further, while
J. Kelsey and B. Schneier proposed a method for finding second preimages
in less than $2^n$, we note that this requires rather long messages
\cite{SchneierSecondPreimage}. However, collision attacks are well within reach
of adversaries, and well past theoretical attacks. The work of
H. Dobbertin \cite{Dobbertin1998}, X. Wang \cite{cryptoeprint:2004:199},
M. Schl{\"a}ffer \cite{Schlaffer2006} and others demonstrate the ease with which
collisions can be found.

@@

We seek to address the problem of evaluating the use cases for a collision,
to determine whether or not a collision affects a particular system. This
paper contributes several techniques to logical cryptanalysis and uses these
techniques to derive metrics of the trustworthiness of hash functions. We claim
the following results are novel in the field:
\begin{itemize}
    \item We define the neighborhood of a class of collisions and show that
        it is frequently non-empty.
    \item We discuss techniques for finding useful collisions within a
        class of collisions.
    \item We propose additional techniques for building new classes of
        collisions and show that collisions in MD4 can be found inductively.
    \item We propose five universal metrics for evaluating the utility of
        a class of collisions.
\end{itemize}
We propose that collision classes of high utility are closer to a second
preimage attack in the amount of flexibility they provide to an attacker,
whereas classes of low utility may not affect all systems which use a
particular hash function.

\textbf{TODO - Finish Organization}
The remainder of this paper is organized as fol
lows. In Section
\ref{Sec:Related} we discuss previous results and show how our research
expands upon prior work. In Section \ref{Sec:Terminology}, we discuss the
terminology and notations we use throughout the remainder of the paper.
In Section \ref{Sec:Intuition}, we give the intuition between measuring
the utility of a collision and motivate why it is useful. In Section
\ref{Sec:Logical}, we present new techniques in logical cryptanalysis to
analyze collisions in hash functions.

@@

This paper draws heavily upon the public cryptanalysis of MD4, the wide
availability of high quality SAT solvers, and previous work in logical
cryptanalysis. In particular, we look at ways that SAT solving can aid
cryptanalysis instead of using cryptanalysis to benchmark SAT solvers. We
then apply them to aid the understanding of attacks against real world
systems, and propose general metrics that target different types of systems.
Towards this, we look at prior work in two main areas: differential and logical
cryptanalysis.

@@

Differential Cryptanalysis has been the major technique behind the discovery
of collisions in cryptographic hash functions. Its importance can be seen
everywhere from X. Wang's attacks on MD4 \cite{Wang2005} to more recent
cryptanalysis of general purpose hash functions such as Murmur3 by
J. Aumasson, D. Bernstein, and M. Bo{\ss}let \cite{murmur3DC}. However, while
differential cryptanalysis plays an important role, techniques for automated
analysis of hash functions are still an emerging area. Further, techniques
are being developed to make hash functions impervious to differential
analysis.

Our work removes the need for finding message modification techniques by
specifying the differential path as part of the SAT formula and letting the
solver find a pair of satisfying messages which follow the differential path
and produce a collision. Further constraints can then be placed upon this
model, such as a chosen prefix or desired start state. We believe performing
these attacks by hand to be difficult, and note that developed message
modification techniques may affect chosen prefixes. Thus, we seek to develop
techniques to replace differential cryptanalysis with logical cryptanalysis.

@@

Logical Cryptanalysis and its encoding as SAT likely started under the work of
F. Massacci in 1999 with his paper titled ``Using walk-SAT and rel-sat for
cryptographic key search''. \cite{Massacci:1999:UWR:1624218.1624261}. However,
much of the early work by F. Massacci was focused on symmetric and asymmetric
ciphers
\cite{Massacci:1999:UWR:1624218.1624261, MASSACCI2000LogicalCA, FIORINI2003101}.
It wasn't until the work of D. Jovanovi{\'{c}} and P. Jani{\v{c}}i{\'{c}}
that exploring collisions in the context of SAT was introduced
\cite{LogicalAnalysis}, and until the work of I. Mironov and L. Zhang that
this was studied for an existing collision class \cite{Mironov2006}.

However, much of the work in this area is focused on benchmarking SAT solvers,
and not on evaluating techniques for using SAT solvers to aid reasoning and
understanding. E. Homsirikamol et al.\ show that SAT solvers can be used to
brute force finding collisions in SHA-3, but with limited utility and limited
to a small number of rounds \cite{Homsirikamol2012}. On the other hand,
advances by V. Nossum in his master's thesis shows that additional work on
encoding is also important: a new 32-bit modular addition circuit produced
shorter run times \cite{nossum_2012}.

We make use of the \textit{bc2cnf} utility by T. Junttila as the basis for
our models \cite{circuits}. This allows us to encode hash functions as generic
structures in the circuit description language and ease the algorithmic
creation of models which build on top of them. We then use CryptoMiniSat5 by
M. Soos to run the models and check for a satisfying witness, often an
example collision \cite{CryptoMiniSat5}. We deviate from evaluating the
performance of SAT solvers or encodings of models by instead focusing on
evaluating the properties of the hash function. We have developed new
techniques for using a SAT solver to inductively build full collisions in MD4
from collisions in reduced-round MD4 and to build addition examples of
collision classes from a single instance of a collision in MD4.

@@

\begin{itemize}
    \item $C_{Wang}$, to be the collision class introduced by X. Wang
        et al. in \cite{Wang2005}.
    \item $C_{Sasaki}$, to be the collision class introduced by Y. Sasaki
        et al. in \cite{Sasaki2007}.
    \item $C_{Dobbertin}$, to be the collision class introduced by
        H. Dobbertin in \cite{Dobbertin1998}.
    \item $C_{Kasselman}$, to be the collision class introduced by
        P. Kasselman in \cite{KasselmanMD4}.
    \item $C_{Schlaffer}$, to be the collision class introduced by
        M. Schl{\"a}ffer in \cite{Schlaffer2006}.
\end{itemize}

@@

\section{ASCII Collisions}

\subsection{Dobbertin}

Default IV

Block 1:

3f474422
624f2b46
683b6c7e
48227d5c
4f2a4f61
206c573e
61622234
227a2633
53482622
706a6727
503e2b3e
7a636c76
7\textbf{b}752620
3f2e5728
236a6376
45654a75

Block 2:

3f474422
624f2b46
683b6c7e
48227d5c
4f2a4f61
206c573e
61622234
227a2633
53482622
706a6727
503e2b3e
7a636c76
7\textbf{c}752620
3f2e5728
236a6376
45654a75

\subsection{Kasselman}

Default IV

Block 1:

7e273957
597e7f4a
7c773d7c
4034376f
69643257
515f5269
6a4d665a
29416b3a
7f377a64
7e232455
6d436c56
26717321
3\textbf{6}65774f
6f2f4421
77787656
26236479

7e273957
597e7f4a
7c773d7c
4034376f
69643257
515f5269
6a4d665a
29416b3a
7f377a64
7e232455
6d436c56
26717321
3\textbf{7}65774f
6f2f4421
77787656
26236479
